Characterizing edges in signed and vector-valued graphs

### 2017-12-13 edge/intro.tex
fix layout of edge|vector chapter

\setlength{\fullpage}{\textwidth+\marginparsep+\marginparwidth}
\begin{table}[h]
  \small
  \setlength{\tabcolsep}{4pt}
  \centering
  \caption{Real attributed graphs that we have collected or are easily accessible. Unfortunately,
    none of them comes with known edge types, which implies the last column is merely speculation.
  \label{tab:edge_apps}}
  {\renewcommand{\arraystretch}{0.95}%
  \begin{tabulary}{\fullpage}{lLLLL}
    \toprule
    {Source} & {Node} & {Edge} & {Node profiles} & {Possible edge meanings} \\
    \midrule
    DBLP & author & citation & keyword count of their papers & scientific topic \\
    Github & developer & following & popularity, number of lines written in various programming languages &
    common hobby, involvement in the same open source project, same
    workplace, famous role model \\
    Amazon & video game products & \enquote{frequently bought together} & price, genre, time of release, popularity and evaluation &
    part of a series, cheap bundle, same genre and level of satisfaction \\
    Wikipedia & article & internal link & category, length, edit history, bag of words &
    generalization, specialization, organisation \\
    Flickr & user & following & activity statistics, groups membership, count of photo tags &
    geography, same tags distribution, famous role model,  \\ 
    Foursquare & venue & \enquote{frequently visited together} & location, popularity, category, the time distribution at which they are visited &
    closeness, part of a \enquote{series} (\eg{} movie theater $\rightarrow$ restaurant $\rightarrow$
    bar $\rightarrow$ club) \\
    IMDB & movie & sharing actors & genre, release date and popularity &
    part of a series, common theme, same time period and genre \\
    BlaBlaCar & driver or passenger & shared ride & experience, location, preferences, rating &
    geography, professional, leisure \\
    \bottomrule
  \end{tabulary}}
\end{table}

\medskip
Besides predicting potentially more than two classes, the other difference with the setting of the
previous chapters is that we now assume that nodes are associated with attributes, which we call
\emph{profiles}. One could argue this makes our framework less general, for attributes are not
always available, or may come with missing values. Another recent finding is that, despite a natural
assumption, the nodes profiles in attributed graphs are not necessarily correlated with known ground
truth communities~\autocite{FortunatoGroundTruth14}, casting further doubt on the usefulness of that
extra information.
This raises the question of why we cannot simply extend the troll/trust method of counting the type
of edges incident to every node and learn a model from such features.
A similar tension between relying solely on always available yet sparse topological information or
leveraging possibly richer attributes information also exists in link
prediction~\autocite{linkPredSurvey16}\stodo{What they actually write is \enquote{Topology-based
techniques are more versatile than attribute-based methods since they are not domain specific} which
sounds a bit unfair: only the attributes are domain specific, not necessarily the methods
themselves}.
% examples of using node attributes for link prediction, taken from that survey:
% https://doi.org/10.1007/s11432-014-5237-y
% - http://dl.acm.org/citation.cfm?id=2124378
% - using graph kernel https://doi.org/10.1016/j.dss.2012.09.019
% - matrix factorization “these latent features (learned from topology) may be combined with
%   optional explicit features for nodes or edges, which yields better performance than using either
%   type of feature exclusively.” https://doi.org/10.1007/978-3-642-23783-6_28
However, the presence of profiles is motivated by the last fundamental difference with the previous
problems. In this chapter, we do not have access to any direct supervision in the form of labels.
Rather than a strict classification problem, we are thus faced with unsupervised clustering, guided
by the information contained in node profiles. Note that contrary to the relation between \pcc{} and
\esp{}, this time the objects we seek to cluster are edges and not nodes. Furthermore, we stress
that the profiles crucially increases the interpretability of our models.

>>>>

\setlength{\fullpage}{179mm}
\begin{table}[h]
\begin{adjustwidth}{-2cm}{}
  \small
  \setlength{\tabcolsep}{4pt}
  \centering
  \caption{Real attributed graphs that we have collected or are easily accessible. Unfortunately,
    none of them comes with known edge types, which implies the last column is merely speculation.
  \label{tab:edge_apps}}
  {\renewcommand{\arraystretch}{0.95}%
  \begin{tabulary}{\fullpage}{lLLLL}
    \toprule
    {Source} & {Node} & {Edge} & {Node profiles} & {Possible edge meanings} \\
    \midrule
    DBLP & author & citation & keyword count of their papers & scientific topic \\
    Github & developer & following & popularity, number of lines written in various programming languages &
    common hobby, involvement in the same open source project, same
    workplace, famous role model \\
    Amazon & video game products & \enquote{frequently bought together} & price, genre, time of release, popularity and evaluation &
    part of a series, cheap bundle, same genre and level of satisfaction \\
    Wikipedia & article & internal link & category, length, edit history, bag of words &
    generalization, specialization, organisation \\
    Flickr & user & following & activity statistics, groups membership, count of photo tags &
    geography, same tags distribution, famous role model,  \\ 
    Foursquare & venue & \enquote{frequently visited together} & location, popularity, category, the time distribution at which they are visited &
    closeness, part of a \enquote{series} (\eg{} movie theater $\rightarrow$ restaurant $\rightarrow$
    bar $\rightarrow$ club) \\
    IMDB & movie & sharing actors & genre, release date and popularity &
    part of a series, common theme, same time period and genre \\
    BlaBlaCar & driver or passenger & shared ride & experience, location, preferences, rating &
    geography, professional, leisure \\
    \bottomrule
  \end{tabulary}}
\end{adjustwidth}
\end{table}

\medskip
Besides predicting potentially more than two classes, the other difference with the setting of the
previous chapters is that we now assume that nodes are associated with attributes, which we call
\emph{profiles}. One could argue this makes our framework less general, for attributes are not
always available, or may come with missing values. Another recent finding is that, despite a natural
assumption, the nodes profiles in attributed graphs are not necessarily correlated with known ground
truth communities~\autocite{FortunatoGroundTruth14}, casting further doubt on the usefulness of that
extra information.
This raises the question of why we cannot simply extend the troll/trust method of counting the type
of edges incident to every node and learn a model from such features.
A similar tension between relying solely on always available yet sparse topological information or
leveraging possibly richer attributes information also exists in link
prediction~\autocite{linkPredSurvey16}. %\stodo{What they actually write is \enquote{Topology-based
%techniques are more versatile than attribute-based methods since they are not domain specific} which
%sounds a bit unfair: only the attributes are domain specific, not necessarily the methods
%themselves}.
% examples of using node attributes for link prediction, taken from that survey:
% https://doi.org/10.1007/s11432-014-5237-y
% - http://dl.acm.org/citation.cfm?id=2124378
% - using graph kernel https://doi.org/10.1016/j.dss.2012.09.019
% - matrix factorization “these latent features (learned from topology) may be combined with
%   optional explicit features for nodes or edges, which yields better performance than using either
%   type of feature exclusively.” https://doi.org/10.1007/978-3-642-23783-6_28
However, the presence of profiles is motivated by the last fundamental difference with the previous
problems. In this chapter, we do not have access to any direct supervision in the form of labels.
Rather than a strict classification problem, we are thus faced with unsupervised clustering, guided
by the information contained in node profiles. Note that contrary to the relation between \pcc{} and
\esp{}, this time the objects we seek to cluster are edges and not nodes. Furthermore, we stress
that the profiles crucially increases the interpretability of our models.

### 2017-12-13 edge/model.tex
typos
fix layout of edge|vector chapter

Stacking all these node profiles in a $|V| \times d$ matrix $X$, our input so far is $(G, X)$. As it
is standard in classification problem, we are also given the number of edge types
$k$.\footnote{Because our problem is essentially unsupervised, assuming that we are given $k$ is a
simplification, on which we elaborate \vpageref{par:edge_choose_k}.} Moreover, recall that we do not
merely want to classify edges into type, but also provide an explanation of why two nodes are
connected. This requires the introduction of three further concepts:
\begin{inparaenum}[1)]
\item how to represent a connection (\ie{} an edge),
\item how to represent an edge's explanation, and
\item how to evaluate which explanation is the best.
\end{inparaenum}
We now present our modelling choice:

\begin{enumerate}[1),nosep,leftmargin=*]
  \item We naturally describe an edge $(u,v)$ through the profiles of its endpoints. More precisely,
    we combine $x_u$ and $x_v$ by an operator $s$ defined as:
    \begin{equation*}
      s(u, v) = x_u \circ x_v = s_{uv} \in \Rbb^d \, ,
    \end{equation*}
    where $\circ$ stands for the Hadamard (or component wise) product. As showed in
    \begin{marginfigure}
      % \small
      \captionsetup{font=small,type=table}
      \captionof{table}{We show the effect of the Hadamard product in combining information from two
        users $u$ and $v$. Instead of actual numbers, in this table we use $+$ to denote a large
        positive value, $-$ to denote a large negative value, $0$ to denote a small absolute value
        and $\star$ to denote an arbitrary value.
      \label{tab:edge_sim}}
      \begin{tabular}{ccc}
        \toprule
        ${x_u}_{;i}$ & ${x_v}_{;i}$ & ${s_{uv}}_{;i}$ \\
        \midrule
        $+$          & $+$          & $+$             \\
        $-$          & $-$          & $+$             \\
        $+$          & $-$          & $-$             \\
        $0$          & $\star$      & $0$             \\
        \bottomrule
      \end{tabular}
    \end{marginfigure}
    \autoref{tab:edge_sim}, for each dimension $i \in \rangesd$, the $s_{uv}$ holds a value
    indicating whether the profiles of $u$ and $v$ agree, disagree or are indifferent along that
    dimension.
  \item The $k$ explanations take the form of bounded norm vectors $\mathcal{D}_k = \{w_1, w_2,
    \ldots, w_k\} \subset \dball$, where $\dball$ is the $\ell_2$-unit ball of $\Rbb^d$. We call such a
    vector $w_\ell$, for $\ell \in \rangesk$, a \emph{direction}. As we shall explain more formally
    later on, those directions are exactly the parameters learned while solving our problem.
  \item Given an edge $(u,v)$, all the $k$ directions $\{w_1, w_2, \ldots, w_k\}$ provide a possible
    explanation of why $u$ and $v$ are connected. We score these explanations by assigning them a
    \emph{goodness} of explanation\footnote{Whereas the name is inspired by the concept of
    \emph{goodness of fit}, there is no formal relation.} defined as:
    \begin{equation*}
      g(s_{uv}, w_\ell) = {s_{uv}}^T w_\ell \, .
    \end{equation*}
    As the $w$ notation hints, a direction can be interpreted as a set of weights. These weights
    indicates to which extent each attribute contributes to the connection, according to that
    direction. For a given set of directions $\mathcal{D}_k$, the classification of edges is
    naturally performed by the following operator $\mathcal{E}$, returning the index of the
    direction with the maximal goodness:
    \begin{equation*}
      \begin{array}[t]{lrcl}
        \mathcal{E} : & E \times \left( \Rbb^d \times \ldots \times \Rbb^d \right)
                      & \longrightarrow & \rangesk \\
                      & (u,v), \mathcal{D}_k & \longmapsto
                      & \argmax_{\ell \in \rangesk} g(s_{uv}, w_\ell)
      \end{array}
    \end{equation*}
    Because it will be clear from context, we lighten the notation and simply write $\mathcal{E}(u,
    v)$ instead of $\mathcal{E}\left((u, v), \mathcal{D}_k \right)$.
\end{enumerate}

@@

For clarity, the previous example of \autoref{fig:edge_exe} has only a handful of nodes and
dimensions, and can therefore reasonably be explained by two directions. But in larger graphs, the
choice of $k$ is a legitimate question.
At one extreme, picking $k=|E|$ renders the topology of the graph
irrelevant, as we can simply exploit the trivial
solution of having a single direction $w_{uv} = \frac{s_{uv}}{||s_{uv}||}$ for every edge $(u,v)$.
More generally, as $k$ increases, we expect the value $\max_{\mathcal{D}_k} \sum_{u,v \in E}
\max_{\ell \in \rangesk} g(s_{uv}, w_{\ell})$ to increase as
well, at the cost of interpretability. There could be a principled approach to finding the
\enquote{best} $k$, based on a information theoretical measure of the complexity of $\mathcal{D}_k$
and the minimum description length principle~\autocite{grunwald2005tutorial}. However, in the
interest of simplicity, in the following we focus on the formulation stated in
\autoref{p:edge_full}, where $k$ is given. In practice, we further constrain $k$ to be small (say
less than \np{10}). Within the multilayer framework, $k$ could also be seen as the number of
underlying layers. To retain a fair level of interpretability, we deem appropriate to have $k$ not
too large.

@@

However,
this does not guarantee that we have assigned the best direction to every edge. We therefore propose
the following heuristic, inspired by the Lloyd algorithm for $k$-means, but where the directions in
$\mathcal{D}_k$ play the role of centroids. Namely, we alternate between two steps:
\begin{enumerate}[(i), nosep]
  \item create a new partition $\{E'_1, \ldots, E'_k\}$ by assigning every edge $(u,v)$ to the
    direction $w$ maximizing its goodness ${s_{uv}}^T w$, that is $E'_\ell = \theset{(u,v) \in E}{
    w_\ell = \argmax_{w_i \in \mathcal{D}_k} {s_{uv}}^T w_i}$, and
  \item update the directions to make them optimal with respect to the edges currently assigned to
    them. Specifically, set $\mathcal{D}_k = \left\{ w_\ell = \frac{s_\ell}{\left\| s_\ell
    \right\|}\right\}_{\ell=1}^k $, where $s_\ell = \sum_{u,v \in E'_\ell} s_{uv}$ is the sum of all
    the similarity vectors of the edge in $E'_\ell$,
\end{enumerate}
until the edge assignment is stable or we reach a maximum number of iterations. By analogy with the
$k$-means algorithm, obtaining convergence guarantees in the general case would be surprising, but in
practice we expect that only a small number $T$ of iterations would be needed. We call this method
\lloyd{} and note it also has a linear complexity of $O(Td|E|)$.

>>>>

Stacking all these node profiles in a $|V| \times d$ matrix $X$, our input so far is $(G, X)$. As it
is standard in classification problem, we are also given the number of edge types
$k$.\footnote{Because our problem is essentially unsupervised, assuming that we are given $k$ is a
simplification, on which we elaborate \vpageref{par:edge_choose_k}.} Moreover, recall that we do not
merely want to classify edges into type, but also provide an explanation of why two nodes are
connected. This requires the introduction of three further concepts:
\begin{enumerate*}[1)]
\item how to represent a connection (\ie{} an edge),
\item how to represent an edge's explanation, and
\item how to evaluate which explanation is the best.
\end{enumerate*}
We now present our modelling choice:

\begin{enumerate}[1),nosep,leftmargin=*]
  \item We naturally describe an edge $(u,v)$ through the profiles of its endpoints. More precisely,
    we combine $x_u$ and $x_v$ by an operator $s$ defined as:
    \begin{equation*}
      s(u, v) = x_u \circ x_v = s_{uv} \in \Rbb^d \, ,
    \end{equation*}
    where $\circ$ stands for the Hadamard (or component wise) product. As showed in
    \autoref{tab:edge_sim}, for each dimension $i \in \rangesd$, the $s_{uv}$ holds a value
    indicating whether the profiles of $u$ and $v$ agree, disagree or are indifferent along that
    dimension.
    \begin{table}[hb]
      \centering
      % \small
      % \captionsetup{font=small,type=table}
      \caption{\small We show the effect of the Hadamard product in combining information from two
        users $u$ and $v$. Instead of actual numbers, in this table we use $+$ to denote a large
        positive value, $-$ to denote a large negative value, $0$ to denote a small absolute value
        and $\star$ to denote an arbitrary value.
      \label{tab:edge_sim}}
      \begin{tabular}{ccc}
        \toprule
        ${x_u}_{;i}$ & ${x_v}_{;i}$ & ${s_{uv}}_{;i}$ \\
        \midrule
        $+$          & $+$          & $+$             \\
        $-$          & $-$          & $+$             \\
        $+$          & $-$          & $-$             \\
        $0$          & $\star$      & $0$             \\
        \bottomrule
      \end{tabular}
    \end{table}
  \item The $k$ explanations take the form of bounded norm vectors $\mathcal{D}_k = \{w_1, w_2,
    \ldots, w_k\} \subset \dball$, where $\dball$ is the $\ell_2$-unit ball of $\Rbb^d$. We call such a
    vector $w_\ell$, for $\ell \in \rangesk$, a \emph{direction}. As we shall explain more formally
    later on, those directions are exactly the parameters learned while solving our problem.
  \item Given an edge $(u,v)$, all the $k$ directions $\{w_1, w_2, \ldots, w_k\}$ provide a possible
    explanation of why $u$ and $v$ are connected. We score these explanations by assigning them a
    \emph{goodness} of explanation\footnote{Whereas the name is inspired by the concept of
    \emph{goodness of fit}, there is no formal relation.} defined as:
    \begin{equation*}
      g(s_{uv}, w_\ell) = {s_{uv}}^T w_\ell \, .
    \end{equation*}
    As the $w$ notation hints, a direction can be interpreted as a set of weights. These weights
    indicates to which extent each attribute contributes to the connection, according to that
    direction. For a given set of directions $\mathcal{D}_k$, the classification of edges is
    naturally performed by the following operator $\mathcal{E}$, returning the index of the
    direction with the maximal goodness:
    \begin{equation*}
      \begin{array}[t]{lrcl}
        \mathcal{E} : & E \times \left( \Rbb^d \times \ldots \times \Rbb^d \right)
                      & \longrightarrow & \rangesk \\
                      & (u,v), \mathcal{D}_k & \longmapsto
                      & \argmax_{\ell \in \rangesk} g(s_{uv}, w_\ell)
      \end{array}
    \end{equation*}
    Because it will be clear from context, we lighten the notation and simply write $\mathcal{E}(u,
    v)$ instead of $\mathcal{E}\left((u, v), \mathcal{D}_k \right)$.
\end{enumerate}

@@

For clarity, the previous example of \autoref{fig:edge_exe} has only a handful of nodes and
dimensions, and can therefore reasonably be explained by two directions. But in larger graphs, the
choice of $k$ is a legitimate question.
At one extreme, picking $k=|E|$ renders the topology of the graph
irrelevant, as we can simply exploit the trivial
solution of having a single direction $w_{uv} = \frac{s_{uv}}{||s_{uv}||}$ for every edge $(u,v)$.
More generally, as $k$ increases, we expect the value \[\max_{\mathcal{D}_k} \sum_{u,v \in E}
\max_{\ell \in \rangesk} g(s_{uv}, w_{\ell})\] to increase as
well, at the cost of interpretability. There could be a principled approach to finding the
\enquote{best} $k$, based on a information theoretical measure of the complexity of $\mathcal{D}_k$
and the minimum description length principle~\autocite{grunwald2005tutorial}. However, in the
interest of simplicity, in the following we focus on the formulation stated in
\autoref{p:edge_full}, where $k$ is given. In practice, we further constrain $k$ to be small (say
less than \np{10}). Within the multilayer framework, $k$ could also be seen as the number of
underlying layers. To retain a fair level of interpretability, we deem appropriate to have $k$ not
too large.

@@

However,
this does not guarantee that we have assigned the best direction to every edge. We therefore propose
the following heuristic, inspired by the Lloyd algorithm for $k$-means, but where the directions in
$\mathcal{D}_k$ play the role of centroids. Namely, we alternate between two steps:
\begin{enumerate}[(i), nosep]
  \item create a new partition $\{E'_1, \ldots, E'_k\}$ by assigning every edge $(u,v)$ to the
    direction $w$ maximizing its goodness ${s_{uv}}^T w$, that is \[E'_\ell = \theset{(u,v) \in E}{
    w_\ell = \argmax_{w_i \in \mathcal{D}_k} {s_{uv}}^T w_i}\]
  \item update the directions to make them optimal with respect to the edges currently assigned to
    them. Specifically, set $\mathcal{D}_k = \left\{ w_\ell = \frac{s_\ell}{\left\| s_\ell
    \right\|}\right\}_{\ell=1}^k $, where $s_\ell = \sum_{u,v \in E'_\ell} s_{uv}$ is the sum of all
    the similarity vectors of the edge in $E'_\ell$,
\end{enumerate}
until the edge assignment is stable or we reach a maximum number of iterations. By analogy with the
$k$-means algorithm, obtaining convergence guarantees in the general case would be surprising, but in
practice we expect that only a small number $T$ of iterations would be needed. We call this method
\lloyd{} and note it also has a linear complexity of $O(Td|E|)$.
